Dark matter halo shape at $z=0$ in the Auriga simulations: radial evolution and alignment with the stellar disk

### 2019-07-19 Document/paper.tex
edit

\title[The shape of dark matter haloes at $z=0$ in Auriga]
{Dark matter halo shape at $z=0$ in the Auriga simulations: radial
  evolution and alignment with the stellar disk}
\author[Jesus Prada,  Jaime E. Forero-Romero, Volker Springel ]{
Jesus Prada,$^{1}$\thanks{E-mail: jd.prada1760@uniandes.edu.co}
Jaime E. Forero-Romero,$^{1}$
Volker Springel$^{2}$
\\
% List of institutions
$^{1}$Departamento de F\'sica, Universidad de los Andes, Cra. 1 No.
18A-10, Edificio Ip, Bogot\'a, Colombia.\\
$^{2}$Heidelberg Institute for Theoretical Studies,
Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany.\\
}

@@

% Abstract of the paper
\begin{abstract}
We present shape measurements of dark matter halos at redshift $z=0$
in a suite of 30 zoom-in simulations from the Auriga project. 
We compare the results in full magnetohydrodynamics against dark
matter only physics. 
We find a strong influence of baryons in making dark matter halos
rounder at all radii compared to its dark matter only counterparts.
This change in triaxility is more pronounced towards the inner
regions of the halo.
At distances $\lesssim 30$ kpc, rounder and coherent dark matter
distributions correlate with extended massive stellar disks and low gas
densities.  
We measure the alignment between the halo and the stellar disk and
find almost perfect alignment at $0.25R_{200}\sim56$ kpc.   
In some cases the alignment between the two components significantly
changes as a function of radius implying that the halo shape twists;
this effect is almost absent in the dark matter only simulations. 
We quantify the twisting effect with the standard deviation of the
angle between the minor axis of the disk and the halo for radii
$\leq 0.5R_{200}$ to show that this effect correlates positively with
extended stellar disks and low gas densities.
In a comparison against observational constraints of dark matter halo
triaxiality for the Milky Way we find that $20\%$ of halos our sample
are consistent with observational results derived from the Pal 5 stream
that favor $b/a\approx 1$ and $c/a>0.77$. Including baryons is a
required element to achieve this level of agreement.
In contrast, none of the simulations (dark matter only and with
baryons) can reproduce the constrains derived from the Saggitarius
stream that favor $b/a\approx 0.97$ and $c/a\approx0.44$. 
\end{abstract}

@@

The Auriga project offers cosmological zoom in simulations of MW-sized 
dark matter halos in a $\Lambda$CDM cosmology. 
This simulations come in two versions: dark matter only and
baryonic physics including magetohydrodynamics (MHD).
A detailed description of the simulations and their
disk properties can be founc in \citep{auriga}, here we summarize its
main features.

The objects in those simulations were selected from a set of 30
isolated halos in the Evolution and Assembly of GaLaxies and their
Environments (EAGLE)  project \citep{Eagle}.   
These halos were randomly selected from a sample of the most isolated
halos at $z=0$ whose virial mass $M_{200}$ was between $10^{12}M_\odot$ and
$2\times 10^{12}M_\odot$. 
The cosmological parameters in these simulatins correspond to
$\Omega_m=0.307$, $\Omega_b=0.048$, $\Omega_\Lambda=0.693$ and a
dimensionless Hubble parameter $h=0.6777$ \citep{2014A&A...571A..16P}

The selected halos were re-simulated at higher resultion by applying a
zoom-in technique with varying physical realism using the moving-mesh AREPO code
that includes gravity, ideal magnetohydrodynamics (MHD), 
phenomenological descriptions for star formation, chemical enrichment
from supernovae and its stellar feedback.  
The simulation also follows the formation and evolution of black holes
together with the Active Galactic Nuclei feedback
\citep{arepo,2013MNRAS.432..176P}.

The 30 zoom-in halos have a dark matter particle mass of $\sim 3\times
10^5$\Msun while the barynic mass resolution is $\sim 5\times 10^4$\Msun.
The softening lenght for gravitational force computation for stellar
particles and high-resolution dark matter particles 
is fixed to be 500 $h^{-1}$ pc in comoving coordinates up to $z=1$,
and 396 pc in physical coordinates afterwards.
The gravitational softening lenght for gas cells changes with the mean
cell radius but is limited to be larger than the stellar softening
lenght and 1850 pc physical. 
In this work all the results that we report at $z=0$ as a function of radius and
the correlations with disk properties correspond to the 30 halos in
the \texttt{Level4} resolution. 
Finally, all  halos described so far have also been simulated at the same
resolutions without MHD using dark matter particles only, we refer to
these halos as the DMO (Dark Matter Only) sample.

>>>>

\title[The shape of dark matter haloes at $z=0$ in Auriga]
{Dark matter halo shape at $z=0$ in the Auriga simulations: radial
  evolution and alignment with the stellar disk}
\author[Jesus Prada,  Jaime E. Forero-Romero, Volker Springel ]{
Jesus Prada,$^{1}$\thanks{E-mail: jd.prada1760@uniandes.edu.co}
Jaime E. Forero-Romero,$^{1}$
Volker Springel$^{2}$
\\
% List of institutions
$^{1}$Departamento de F\'sica, Universidad de los Andes, Cra. 1 No.
18A-10, Edificio Ip, Bogot\'a, Colombia\\
$^{2}$Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzschild-Str. 1, D-85741 Garching, Germany\\
}

@@

% Abstract of the paper
\begin{abstract}
We present shape measurements of dark matter halos at redshift $z=0$
in a suite of 30 zoom-in simulations from the Auriga project. 
We compare the results in full magnetohydrodynamics against dark
matter only physics. 
We find a strong influence of baryons in making dark matter halos
rounder at all radii compared to its dark matter only counterparts.
At distances $\lesssim 30$ kpc, rounder and coherent dark matter
distributions correlate with extended massive stellar disks and low gas
densities.  
We measure the alignment between the halo and the stellar disk and
find almost perfect alignment at $0.25R_{200}\sim56$ kpc.   
In some cases the alignment between the two components significantly
changes as a function of radius implying that the halo shape twists;
this effect correlates with extended stellar disks and
low gas densities and is almost absent in the dark matter only
simulations.  
In a comparison against observational constraints  we find that $20\%$
of halos are consistent with observational results derived
from the Pal 5 stream that favor $b/a\approx 1$ and
$c/a>0.77$. Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (dark matter
only and with baryons) can reproduce the constrains derived from the
Saggitarius stream that favor $b/a\approx 0.97$ and $c/a\approx0.44$.
\end{abstract}

@@

The Auriga project offers cosmological zoom in simulations of MW-sized 
dark matter halos in a $\Lambda$CDM cosmology. 
This simulations come in two versions: dark matter only (DMO) and
baryonic physics including magetohydrodynamics (MHD).
A detailed description of the simulations and their disk properties
can be founc in \citep{auriga}, here we summarize its main features.

The objects in the simulations were selected from a set of 30
isolated halos in the Evolution and Assembly of GaLaxies and their
Environments (EAGLE)  project \citep{Eagle}.   
These halos were randomly selected from a sample of the most isolated
halos at $z=0$ whose virial mass $M_{200}$ was between $10^{12}M_\odot$ and
$2\times 10^{12}M_\odot$. 
The cosmological parameters in these simulatins correspond to
$\Omega_m=0.307$, $\Omega_b=0.048$, $\Omega_\Lambda=0.693$ and a
dimensionless Hubble parameter $h=0.6777$ \citep{2014A&A...571A..16P}

The selected halos were re-simulated at higher resultion by applying a
zoom-in technique with varying physical realism using the moving-mesh AREPO code
that includes gravity, ideal magnetohydrodynamics,  phenomenological
descriptions for star formation, chemical enrichment from supernovae
and its stellar feedback.   
The simulation also follows the formation and evolution of black holes
together with the Active Galactic Nuclei feedback
\citep{arepo,2013MNRAS.432..176P}.

The 30 zoom-in halos have a dark matter particle mass of $\sim 3\times
10^5$\Msun while the barynic mass resolution is $\sim 5\times 10^4$\Msun.
The softening lenght for gravitational force computation for stellar
particles and high-resolution dark matter particles 
is fixed to be 500 $h^{-1}$ pc in comoving coordinates up to $z=1$,
and 396 pc in physical coordinates afterwards.
The gravitational softening lenght for gas cells changes with the mean
cell radius but is limited to be larger than the stellar softening
lenght and 1850 pc physical. 
From now on we refer to the haloes simulated with baryonic physics as the
MHD sample and to the haloes simulated with dark matter only as the
DMO sample.
